---
title: "Assignment 3 - Navid Farzanpay"
author: "Navid Farzanpay"
date: "2026-02-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```
```{r}
library(caret)
library(e1071)
bank<-read.csv("UniversalBank-1.csv")
```
## Prepare data
```{r}
bank_small<-bank[,c("Personal.Loan","Online","CreditCard")]
bank_small$Personal.Loan<-factor(bank_small$Personal.Loan)
bank_small$Online<-factor(bank_small$Online)
bank_small$CreditCard<-factor(bank_small$CreditCard)
```
## Partition into training and validation sets
```{r}
set.seed(123)
idx_train<-createDataPartition(bank_small$Personal.Loan,p=0.60,list = FALSE)
train_df<-bank_small[idx_train,]
valid_df<-bank_small[-idx_train,]
```
## Part A
```{r}
pivot_A<-table(CC=train_df$CreditCard, Loan=train_df$Personal.Loan, Online=train_df$Online)
pivot_A
```
## Part B
```{r}
prob_pivot<-51/(467+51)
prob_pivot
```
### Answer to part B
The probability that a customer with CC=1 and Online=1 accepts the loan is 9.85%

## Part C
```{r}
#Loan as a function of online
pivot_online<-table(Loan=train_df$Personal.Loan, Online=train_df$Online)
pivot_online
#Loan as a function of credit card
pivot_cc<-table(Loan=train_df$Personal.Loan, CC=train_df$CreditCard)
pivot_cc
```
## Part D
```{r}
# Totals
total_loan1<-sum(train_df$Personal.Loan==1)
total_loan0<-sum(train_df$Personal.Loan==0)
total_n<-nrow(train_df)
# i
p_cc1_given_loan1<-sum(train_df$CreditCard==1&train_df$Personal.Loan==1)/total_loan1
# ii
p_online1_given_loan1<-sum(train_df$Online==1&train_df$Personal.Loan==1)/total_loan1
# iii
p_loan1<-total_loan1/total_n
# iv
p_cc1_given_loan0<-sum(train_df$CreditCard==1&train_df$Personal.Loan==0)/total_loan0
# v
p_online1_given_loan0<-sum(train_df$Online==1&train_df$Personal.Loan== 0)/total_loan0
# vi
p_loan0<-total_loan0/total_n
p_cc1_given_loan1
p_online1_given_loan1
p_loan1
p_cc1_given_loan0
p_online1_given_loan0
p_loan0
```
## Part E
```{r}
#Numerator for loan=1
num_loan1<-p_cc1_given_loan1*p_online1_given_loan1*p_loan1
# Numerator for loan=0
num_loan0<-p_cc1_given_loan0*p_online1_given_loan0*p_loan0
# Naive Bayes probability
prob_nb<-num_loan1/(num_loan1+num_loan0)
prob_nb
```
### The naive bayes estimate is P(loan=1|cc=1,online=1)=0.1001 (about 10.01%)
## Part F
From the pivot table, the probability is 9.85%
From the naive bayes calculation, the probabilty is 10.01%
The pivot table value is more accurate because it is based directly on the actual counts in the training data, while the naive bayes value is slightly different because it assumes that online and creditcard are independent given loan

## Part G
```{r}
#Fit naive bayes model
nb_model<-naiveBayes(Personal.Loan~CreditCard+Online,data=train_df)
nb_model
```
```{r}
# Create a new customer with cc=1 and online=1
new_customer<-data.frame(CreditCard=factor(1,levels=levels(train_df$CreditCard)),
                         Online=factor(1,levels = levels(train_df$Online)))
#Get predicted probabilities
predict(nb_model,new_customer,type="raw")
```
The naive bayes model gives a probability of 10.01% for loan=1 when cc=1 and online=1. This matches the value computed manually in Part E, confirming that the manual naive bayes calculation was correct.
