---
title: "Assignment 1"
author: "Navid Farzanpay"
date: "2026-01-29"
output: html_document
---
## Dataset

For this assignment I used the U.S. Department of Transportation Consumer Airfare Report, Table 2 (Top 1,000 City-Pair Markets). The dataset includes city-level aggregated market data such as passenger counts, average fares, yield, and distance.

## Load data

```{r}
airfare <- read.csv("Consumer_Airfare_Report__Table_2_-_Top_1_000_City-Pair_Markets.csv")
dim(airfare)
```
The dataset contains 8,509 observations and 16 variables, representing aggregated city-pair airfare market data.

## Descriptive statistics

```{r}
summary(airfare$cur_fare)
```

This section summarizes the distribution of average airfare across 
city-pair markets.

## Categorical variable: City frequency

```{r}
table(airfare$city)
```

This table shows how frequently each city appears in the top 1,000 
city-pair markets.

## Histogram of average fares

```{r}
hist(airfare$cur_fare,
     main = "Distribution of Average Airfares",
     xlab = "Average Fare (USD)")
```

The histogram shows that most average fares fall within a central range, with fewer markets at very low or very high prices.

